---
title: "Priors Aren't the Problem — They're the Point"
subtitle: "A simulation study on historical control borrowing in Bayesian clinical trials"
author: "Maggie Qian"
date: "2026-03-12"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
    fig-dpi: 180
  pdf:
    toc: true
    fig-width: 10
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Overview

A common claim in Bayesian vs. Frequentist debates is that with an uninformative
prior the two frameworks give identical answers — and therefore the prior doesn't
matter. This is technically true in the limit, but it misses the point entirely.

**The value of an informative prior on the control arm isn't that it changes
your conclusion. It's that borrowing from historical control data lets you reach
the same target power at a lower per-arm sample size — at a controlled,
calibrated Type I error rate.**

This simulation study uses a **historical control borrowing** setup: we place an
informative Beta prior on the control arm only (encoding different amounts of
historical data at the same prior mean), while keeping the treatment prior flat.
This is a common and well-studied use case for informative priors in clinical
trials, and one that FDA's January 2026 Bayesian guidance addresses directly.

Three claims are examined:

1. At a **fixed, calibrated Type I error**, borrowing historical control data
   increases power for the same per-arm sample size
2. **Prior misspecification** shifts operating tradeoffs: a pessimistic prior
   is conservative and loses power; an optimistic prior requires stricter
   calibration and may behave less robustly under prior–data conflict
3. The **likelihood eventually dominates** any prior — the self-correcting
   property of Bayesian updating

---

## Setup

```{r setup}
library(tidyverse)
library(patchwork)

set.seed(42)

# ── Global parameters ────────────────────────────────────────────────────────
TRUE_CTRL    <- 0.20
TRUE_TRT     <- 0.35
N_SIMS       <- 2000   # simulations per (prior, N) cell; increase to 5000-10000 for publication-quality smoothness
N_SIMS_BANDS <- 500    # simulations for Part 3 convergence bands
TARGET_ALPHA <- 0.05   # target one-sided Type I error for calibration
N_SEQ        <- c(30, 50, 75, 100, 125, 150, 175, 200, 250, 300)

# ── Theme ─────────────────────────────────────────────────────────────────────
theme_eb <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13),
      plot.subtitle    = element_text(color = "grey40", size = 10),
      plot.caption     = element_text(color = "grey50", size = 8.5),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom",
      legend.title     = element_blank()
    )
}

prior_colors <- c(
  "Minimal ESS Beta(1,4)" = "#9ecae1",
  "Weak Beta(4,16)"       = "#4292c6",
  "Strong Beta(10,40)"    = "#084594"
)

robust_colors <- c(
  "Correct \u03bc=0.20"     = "#2ca25f",
  "Pessimistic \u03bc=0.30" = "#de2d26",
  "Optimistic \u03bc=0.08"  = "#f16913"
)
```

---

## Helper functions

### Analytical posterior probability

For two independent Beta posteriors, P(θ_t > θ_c) can be computed via
one-dimensional numerical integration — no Monte Carlo noise in the inner loop.

$$
P(\theta_t > \theta_c \mid \text{data}) =
\int_0^1 \bigl[1 - F_{\theta_t}(y)\bigr] \, f_{\theta_c}(y) \, dy
$$

where $F_{\theta_t}$ is the Beta CDF with posterior parameters
$(a_t + s_t,\; b_t + n_t - s_t)$ and $f_{\theta_c}$ is the Beta density with
$(a_c + s_c,\; b_c + n_c - s_c)$.

```{r helpers}
# Exact (deterministic) P(theta_t > theta_c) via numerical integration
prob_beats_exact <- function(s_t, n_t, s_c, n_c, a_t, b_t, a_c, b_c) {
  at <- a_t + s_t;  bt <- b_t + n_t - s_t
  ac <- a_c + s_c;  bc <- b_c + n_c - s_c
  integrate(
    function(y) (1 - pbeta(y, at, bt)) * dbeta(y, ac, bc),
    lower = 0, upper = 1, rel.tol = 1e-5
  )$value
}

# Simulate N_SIMS P(trt > ctrl) values under a specified true_trt
# (set true_trt = true_ctrl for H0 / calibration)
simulate_probs <- function(a_c, b_c, n, n_sims,
                            true_ctrl, true_trt,
                            a_t = 1, b_t = 1) {
  s_c <- rbinom(n_sims, n, true_ctrl)
  s_t <- rbinom(n_sims, n, true_trt)
  map_dbl(seq_len(n_sims), \(i)
    prob_beats_exact(s_t[i], n, s_c[i], n, a_t, b_t, a_c, b_c)
  )
}
```

### Design calibration

For each (prior, N), we find the threshold τ such that the empirical Type I error
≈ `r TARGET_ALPHA` under H0. Power is then evaluated at that same τ.

```{r calibrate}
# Returns a named list: threshold, type1_error, power
calibrated_design <- function(a_c, b_c, n, n_sims,
                               true_ctrl, true_trt,
                               target_alpha = TARGET_ALPHA,
                               a_t = 1, b_t = 1) {

  h0_probs <- simulate_probs(a_c, b_c, n, n_sims, true_ctrl, true_ctrl, a_t, b_t)
  tau       <- quantile(h0_probs, 1 - target_alpha)   # calibrated threshold

  h1_probs  <- simulate_probs(a_c, b_c, n, n_sims, true_ctrl, true_trt,  a_t, b_t)

  list(
    threshold  = as.numeric(tau),
    type1_err  = mean(h0_probs >= tau),
    power      = mean(h1_probs >= tau)
  )
}
```

---

## Part 1: Power gain from historical control borrowing

Three control-arm priors, **all centered at the true control rate of 20%**,
varying only in informativeness (ESS).

| Prior | ESS | Encoding |
|-------|-----|----------|
| Beta(1,4) | 5 | ~5 historical patients |
| Beta(4,16) | 20 | ~20 historical patients |
| Beta(10,40) | 50 | ~50 historical patients |

The treatment prior is flat Beta(1,1) throughout. Each design is calibrated
to 5% Type I error at every sample size before comparing power.

```{r part1-sim}
priors_p1 <- list(
  "Minimal ESS Beta(1,4)" = c(a = 1,  b = 4),
  "Weak Beta(4,16)"       = c(a = 4,  b = 16),
  "Strong Beta(10,40)"    = c(a = 10, b = 40)
)

# Sanity check prior means (all should be TRUE_CTRL = 0.20)
imap_dfr(priors_p1, \(pr, label) tibble(
  prior = label,
  mean  = pr["a"] / (pr["a"] + pr["b"]),
  ess   = pr["a"] + pr["b"]
))
```

```{r part1-power, cache=TRUE, cache.extra=1}
power_results <- imap_dfr(priors_p1, \(pr, label) {
  message("Part 1 | ", label)
  map_dfr(N_SEQ, \(n) {
    res <- calibrated_design(pr["a"], pr["b"], n, N_SIMS, TRUE_CTRL, TRUE_TRT)
    tibble(n = n, power = res$power, type1 = res$type1_err,
           tau = res$threshold, prior = label)
  })
})

# N at first ≥ 80% calibrated power
# Note: if a curve never crosses 80%, this will return an empty tibble for that prior
n80 <- power_results |>
  group_by(prior) |>
  filter(power >= 0.80) |>
  slice(1) |>
  select(prior, n80 = n, power, tau)

n80
```

```{r fig1, fig.cap="Calibrated power curves: all designs hold Type I error at 5%."}
ggplot(power_results, aes(n, power, color = prior)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey50") +
  geom_vline(
    data = n80,
    aes(xintercept = n80, color = prior),
    linetype = "dotted", linewidth = 0.9
  ) +
  annotate("text", x = 305, y = 0.815, label = "80% power",
           color = "grey50", size = 3.2, hjust = 1) +
  scale_color_manual(values = prior_colors) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_x_continuous(breaks = N_SEQ) +
  labs(
    title    = "Figure 1 — Historical Control Borrowing Increases Power at Fixed Type I Error",
    subtitle = glue::glue(
      "True control: {scales::percent(TRUE_CTRL)}  |  ",
      "True treatment: {scales::percent(TRUE_TRT)}  |  ",
      "All designs calibrated to {scales::percent(TARGET_ALPHA)} Type I error"
    ),
    x       = "Patients per arm",
    y       = "Calibrated power",
    caption = paste(
      "All three priors centered at 20% (true control rate); ESS varies.",
      "Calibrated threshold \u03c4 estimated separately for each (prior, N) via simulation quantile.",
      "Dotted verticals mark first N \u2265 80% power. Treatment prior: flat Beta(1,1) throughout.",
      sep = "\n"
    )
  ) +
  theme_eb()
```

```{r fig1b, fig.cap="Empirical Type I error after calibration (should be near 5% throughout)."}
ggplot(power_results, aes(n, type1, color = prior)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_hline(yintercept = TARGET_ALPHA, linetype = "dashed", color = "grey50") +
  annotate("text", x = 305, y = TARGET_ALPHA + 0.008, label = "5% target",
           color = "grey50", size = 3.2, hjust = 1) +
  scale_color_manual(values = prior_colors) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.15)) +
  scale_x_continuous(breaks = N_SEQ) +
  labs(
    title    = "Figure 1b — Empirical Type I Error After Design Calibration",
    subtitle = "Deviations from 5% reflect simulation noise in the calibration step",
    x        = "Patients per arm",
    y        = "Empirical Type I error"
  ) +
  theme_eb()
```

> **Key result:** At a **common 5% Type I error**, the strong prior (ESS = 50)
reaches 80% power at
N = `r n80 |> dplyr::filter(prior == "Strong Beta(10,40)") |> dplyr::pull(n80)`/arm, versus
N = `r n80 |> dplyr::filter(prior == "Minimal ESS Beta(1,4)") |> dplyr::pull(n80)`/arm for the
minimal-ESS prior — saving
`r (n80 |> dplyr::filter(prior == "Minimal ESS Beta(1,4)") |> dplyr::pull(n80) - n80 |> dplyr::filter(prior == "Strong Beta(10,40)") |> dplyr::pull(n80)) * 2`
patients total. The power gain is a clean consequence of ESS, not of shifting
the prior mean.

---

## Part 2: Robustness to prior misspecification

What happens when the historical data was misleading? We hold ESS ≈ 50 fixed
and vary the prior mean, then compare **calibrated** power and the empirical
Type I error under each scenario's own calibrated threshold.

Because calibration is done per-scenario under that scenario's own null,
any remaining deviation from 5% in the figures below reflects finite simulation
error in the calibration step, not a fundamental failure of the approach. The
more meaningful operating differences show up in the calibrated threshold itself
and in power: a misspecified prior changes what threshold the design requires,
and can shift power substantially depending on the direction of misspecification.

| Scenario | Prior | Prior Mean | True Control |
|----------|-------|-----------|--------------|
| Correct | Beta(10,40) | 20% | 20% |
| Pessimistic | Beta(15,35) | 30% | 20% |
| Optimistic | Beta(4,46) | 8% | 20% |

```{r part2-sim, cache=TRUE}
rob_priors <- list(
  "Correct \u03bc=0.20"     = c(a = 10, b = 40),
  "Pessimistic \u03bc=0.30" = c(a = 15, b = 35),
  "Optimistic \u03bc=0.08"  = c(a = 4,  b = 46)
)

robustness_results <- imap_dfr(rob_priors, \(pr, label) {
  message("Part 2 | ", label)
  map_dfr(N_SEQ, \(n) {
    res <- calibrated_design(pr["a"], pr["b"], n, N_SIMS, TRUE_CTRL, TRUE_TRT)
    tibble(n = n, power = res$power, type1 = res$type1_err,
           tau = res$threshold, scenario = label)
  })
})

# Key numbers at N = 150
robustness_results |>
  filter(n == 150) |>
  select(scenario, tau, type1, power) |>
  mutate(across(c(type1, power), \(x) scales::percent(x, accuracy = 0.1)))
```

```{r fig2, fig.cap="Calibrated power (left) and residual Type I error (right) under prior misspecification."}
p2a <- ggplot(robustness_results, aes(n, power, color = scenario)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey60") +
  annotate("text", x = 305, y = 0.815, label = "80%",
           color = "grey50", size = 3, hjust = 1) +
  scale_color_manual(values = robust_colors) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_x_continuous(breaks = c(50, 100, 150, 200, 250, 300)) +
  labs(title = "Calibrated Power", x = "Patients per arm", y = "Power") +
  theme_eb() +
  theme(legend.position = "right")

p2b <- ggplot(robustness_results, aes(n, type1, color = scenario)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_hline(yintercept = TARGET_ALPHA, linetype = "dashed", color = "grey60") +
  annotate("text", x = 305, y = TARGET_ALPHA + 0.008, label = "5% target",
           color = "grey50", size = 3, hjust = 1) +
  scale_color_manual(values = robust_colors) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.20)) +
  scale_x_continuous(breaks = c(50, 100, 150, 200, 250, 300)) +
  labs(title = "Residual Type I Error After Calibration",
       x = "Patients per arm", y = "Type I error") +
  theme_eb() +
  theme(legend.position = "right")

(p2a / p2b) +
  plot_annotation(
    title    = "Figure 2 — Prior Misspecification Shifts Operating Tradeoffs",
    subtitle = "ESS \u2248 50 throughout. True control rate: 20%. Threshold calibrated per scenario.",
    caption  = paste(
      "Pessimistic prior (μ=0.30): overestimates control rate \u2192 conservative, lower power.",
      "Optimistic prior (μ=0.08): underestimates control rate \u2192 requires stricter threshold, less robust under drift.",
      "Remaining deviations from 5% Type I error reflect simulation noise in calibration.",
      sep = "\n"
    ),
    theme = theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(color = "grey40", size = 10),
      plot.caption  = element_text(color = "grey50", size = 8.5)
    )
  )
```

> **Key result:** The main effect of misspecification shows up in the calibrated
threshold τ and in power, not in persistent Type I error inflation — which is
largely absorbed by the per-scenario calibration. A pessimistic prior is
conservative: it requires a lower threshold and loses power. An optimistic prior
requires a stricter calibrated threshold and is more sensitive to prior–data
conflict; under real-world drift between historical and current controls, that
lack of robustness becomes more consequential. Any remaining
deviation from 5% in Figure 2 (right panel) is primarily simulation noise.

---

## Part 3: The self-correcting property

The posterior mean of the control arm converges toward the true rate as data
accumulate. To avoid an artifact of a single lucky or unlucky simulation path,
we run `r N_SIMS_BANDS` independent realisations and report the **median** plus
**50% and 95% simulation quantile bands** across paths.

```{r part3-sim, cache=TRUE}
run_convergence_paths <- function(a, b, n_max, true_ctrl, n_paths) {
  n_trace <- seq(5, n_max, by = 5)
  # matrix: rows = trace points, cols = paths
  mat <- replicate(n_paths, {
    obs <- rbinom(n_max, 1, true_ctrl)
    cumsum_obs <- cumsum(obs)[n_trace]
    (a + cumsum_obs) / (a + b + n_trace)
  })
  # Summarise across paths
  map_dfr(seq_along(n_trace), \(i) tibble(
    n      = n_trace[i],
    median = median(mat[i, ]),
    lo95   = quantile(mat[i, ], 0.025),
    hi95   = quantile(mat[i, ], 0.975),
    lo50   = quantile(mat[i, ], 0.25),
    hi50   = quantile(mat[i, ], 0.75)
  ))
}

convergence_results <- imap_dfr(rob_priors, \(pr, label) {
  run_convergence_paths(pr["a"], pr["b"], 300, TRUE_CTRL, N_SIMS_BANDS) |>
    mutate(scenario = label)
})
```

```{r fig3, fig.cap="Median posterior mean ± bands across 500 simulated control arms."}
ggplot(convergence_results, aes(n, color = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = lo95, ymax = hi95), alpha = 0.10, color = NA) +
  geom_ribbon(aes(ymin = lo50, ymax = hi50), alpha = 0.20, color = NA) +
  geom_line(aes(y = median), linewidth = 1.2) +
  geom_hline(yintercept = TRUE_CTRL, linetype = "dashed",
             color = "grey30", linewidth = 0.9) +
  annotate("text", x = 305, y = TRUE_CTRL -0.05,
           label = paste0("True rate = ", scales::percent(TRUE_CTRL)),
           color = "grey30", size = 3.2, hjust = 1) +
  scale_color_manual(values = robust_colors) +
  scale_fill_manual(values  = robust_colors) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0.02, 0.42)) +
  scale_x_continuous(breaks = c(50, 100, 150, 200, 250, 300)) +
  labs(
    title    = "Figure 3 — The Likelihood Eventually Overwhelms Any Prior",
    subtitle = paste(
      "Median posterior mean of control arm over {N_SIMS_BANDS} simulations",
      "Bands: 50% (dark) and 95% (light) empirical quantile bands across simulated paths",
      "ESS \u2248 50 across all priors",
      sep = "\n"
    ),
    x       = "Cumulative patients observed (control arm)",
    y       = "Posterior mean",
    caption = paste(
      "Pessimistic prior starts high; optimistic starts low.",
      "Medians approach the true rate roughly in the N = 100\u2013150 range (visually approximate),",
      "the self-correcting property of Bayesian updating.",
      sep = "\n"
    )
  ) +
  theme_eb()
```

> **Key result:** The median of all three posteriors converges to the true 20%
rate by approximately N = 100–150. The 95% bands narrow predictably with sample
size. This confirms the self-correcting claim on a stable, replication-robust
basis rather than a single path.

---

## Summary

```{r summary-table}
# Join ESS onto n80, then derive baseline from the lowest-ESS design
n80_annotated <- n80 |>
  dplyr::left_join(
    imap_dfr(priors_p1, \(pr, label) tibble(
      prior = label,
      ess   = pr["a"] + pr["b"],
      mean  = scales::percent(pr["a"] / (pr["a"] + pr["b"]))
    )),
    by = "prior"
  )

flat_n80 <- n80_annotated |>
  dplyr::ungroup() |>
  dplyr::arrange(ess) |>
  dplyr::slice(1) |>
  dplyr::pull(n80)

n80_annotated |>
  dplyr::transmute(
    Prior                    = prior,
    ESS                      = ess,
    `Prior Mean`             = mean,
    `N at 80% Power`         = n80,
    `Patients Saved vs Minimal ESS` = (flat_n80 - n80) * 2
  ) |>
  knitr::kable(align = "lrrcc")
```

**Bottom line:** Informative priors on the control arm aren't a way to cheat
the data. They're a mechanism for putting existing historical knowledge to work —
at a controlled, calibrated Type I error rate. The sample size savings are real,
the operating consequences of misspecification are quantifiable and manageable,
and with enough accumulating data, the likelihood pulls the posterior toward the truth even when the prior is misspecified. FDA's January 2026
Bayesian guidance addresses this framework explicitly.

---

## Session info

```{r session}
sessionInfo()
```
